---
wp_title: Functions
wp_description: A look at mathematical functions
wp_post_type: post
wp_categories: Algebra I, test 2, test 3
wp_mt_keywords: algebra i, math functions
wp_draft: TRUE
---

```{r libraries, include = FALSE}
library(ggplot2)
```

##Introduction to functions

###What is a function?

[KhanAcademy](https://www.khanacademy.org/math/algebra/algebra-functions/intro-to-functions/v/what-is-a-function)

A function is a formula that will take an input and produce a given output. 

######Example: 

$$  f(x) = \begin{cases} \text{if } x = even, x^2 \\ \text{otherwise}, x + 5 \end{cases}  $$

We're asking here if $x$ is an even number then return the square root of x. Otherwise, return $x + 5$.

So, if $x = 2$ then since $2$ is an even number the function returns the value of $2^2$ or $4$.  If $x = 3$ then since $3$ is odd the function would return $3 + 5$ or $8$.

######Example:

$$ h(a) = \lceil a \rceil $$

Here, the formula asks us to take input $a$ and return the next largest integer. If we pass $a = 2$ the formula returns $3$. If we pass $a = 3$, the formula returns $4$.

######Example:

$$ y = x + 1 $$

This equation is the same as saying $y = f(x) = x + 1$.  In this example, $f(0) = 1$, $f(1) = 2$, etc...

A function produces only one possible output. It is not a function if you can have multiple outputs.

##Evaluating functions

The function $f(x)$ is defined as $f(x) = 49 - x^2$. 

Find the value of $f(5)$. Here, $x = 5$ so the functin would calculate $49 - 5^2$ which is $49 - 25$ which returns 24. Therefore, $f(5) = 24$. 

The function $f(x)$ is graphed below. Find $f(-1)$. 

```{r}
set.seed(24880)
m <- data.frame(x = seq(-10, 10, by = 1), 
                y = sample(4:8, 21, replace = TRUE))

ggplot(data = m, aes(x = x, y = y)) + 
    geom_line() + 
    expand_limits(y = c(-10, 10)) + 
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) +
    scale_y_continuous(breaks = seq(-10, 10, 1)) + 
    scale_x_continuous(breaks = seq(-10, 10, 1))
```

We're looking for the output of $f(x)$ when $x = -1$. By looking at the graph, we find where $x = -1$, then $y = 4$. 

```{r}
ggplot(data = m, aes(x = x, y = y)) + 
    geom_line() + 
    expand_limits(y = c(-10, 10)) + 
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) + 
    geom_point(x = -1, y = 4) + 
    geom_text(label = "y = f(x)", x = -2, y = 3) + 
    scale_y_continuous(breaks = seq(-10, 10, 1)) + 
    scale_x_continuous(breaks = seq(-10, 10, 1))
```

###Evaluate an expression that contains function notation

```{r}
set.seed(1)
m <- data.frame(x = rep(seq(-9, 10, by = 1), times = 1, each = 1), 
                y = c(sample(0:10, 20, replace = TRUE), sample(-10:0, 20, replace = TRUE)), 
                z = rep(c("f", "g"), times = 1, each = 20))

ggplot(data = m, aes(x = x, y = y, colour = z)) + 
    geom_line() + 
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) + 
    scale_y_continuous(breaks = seq(-10, 10, 1)) + 
    scale_x_continuous(breaks = seq(-10, 10, 1)) + 
    theme(legend.title = element_blank()) + 
    scale_colour_discrete(labels = c("f(x)", "g(x)"))
```

What if we are asked to find $-2 * f(-6) + g(1)$ on the above plot? For this, I'll assume that $x = -6$ for $f(-6)$ and $x = 1$ for $g(1)$. 

To determine the expression we first identify the $y$ value of $f(-6)$ which is 9. Then I want to determine the $y$ value of $g(1)$ which is -5. 

```{r}
ggplot(data = m, aes(x = x, y = y, colour = z)) + 
    geom_line() + 
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) + 
    scale_y_continuous(breaks = seq(-10, 10, 1)) + 
    scale_x_continuous(breaks = seq(-10, 10, 1)) + 
    theme(legend.title = element_blank()) + 
    scale_colour_discrete(labels = c("f(x)", "g(x)")) + 
    geom_point(x = -6, y = 9, colour = "red") + 
    geom_text(label = "f(-6)", x = -6, y = 10, colour = "red") + 
    geom_point(x = 1, y = -5, colour = "blue") + 
    geom_text(label = "g(1)", x = 1.75, y = -5.5, colour = "blue")
```

Now we simply plug in the output values:

$$-2 * f(-6) + g(1)$$
$$-2 * 9 + -5$$
$$-18 - 5 = -23$$

The function $f$ is defined as follows:

$$f(t) = -2t + 5$$

What is the input value for which $f(t) = 13$?

$$f(t) = -2t + 5 = 13$$
$$-2t = (13-5)$$
$$-2t = 8$$
$$t = \frac{8}{-2}$$
$$t = -4$$
$$f(-4) = -2t + 5$$

####The Problem

**What is the input value for which the output of $h(t) = 4t + 20$ is equal to 4?**

#####The Strategy

We need to sovle an equation with the function's formula $rt + 20$ on one side and the given output $4$ on the other side.

######Solving the equation
$$4t + 20 = 4$$
$$4t = -16$$
$$t = -4$$

Therfore, $h(-4) = 4.

**The answer**
$t = -4$

###How to find two function inputs with the same output given graph

```{r}
m <- data.frame(x = c(seq(-10, 10)), 
                y = c(-2, 0, 1, 2, 2, 4, 3, 2, 1, 3, 2, 1, 2, 3, 4, 5, 5, 5, 6, 6, 5))

ggplot(data = m, aes(x = x, y = y)) + 
    geom_line() + 
    expand_limits(y = c(-10, 10)) + 
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) + 
    scale_y_continuous(breaks = seq(-10, 10, 1)) + 
    scale_x_continuous(breaks = seq(-10, 10, 1))
```

The graph of the function $f$ is show above. 

**What is the input value other than $-5$ for which $f(x) = f(-5)$?

We identify that $f(-5) = 4$ and we look for another $x$ value that also gives us an output of $4$. 

 ```{r}
 ggplot(data = m, aes(x = x, y = y)) + 
    geom_line() + 
    expand_limits(y = c(-10, 10)) + 
    geom_vline(xintercept = 0) + 
    geom_hline(yintercept = 0) + 
    scale_y_continuous(breaks = seq(-10, 10, 1)) + 
    scale_x_continuous(breaks = seq(-10, 10, 1)) + 
    geom_point(x = -5, y = 4, colour = "red") + 
    geom_text(label = "f(-5)", x = -5, y = 5, colour = "red") + 
    geom_hline(yintercept = 4, colour = "red")
```

We can see on that graph above by the horizontal red line that $f(x)$ doesn't equal 4 again until $x = 4$. Therefore, our solution would be 4.

###Generate input-output pairs of a function

We are asked:

> Pick any three pairs of corresponding input and output values of the following function, and fill the table accordingly. If necessary, round your answers to the nearest 0.1.

$$g(x) = -3(x-3) + 1$$

$$\begin{array}{c|c|c|c}
    x    & i & i & i \\ \hline
    g(x) & j & j & j \\
\end{array}$$

So, we are asked to populate the $i$ and $j$ in the table based on the function provided. But, where to start? 

Well, simply put, we can start wherever we want - at least as far as $x$ is concerned. So, why not just start with $x = 0$? In fact, I'll make $x$ a range between $-1$ and $1$ and see how that works.

$$\begin{array}{c|c|c|c}
    x    & -1 & 0 & 1 \\ \hline
    g(x) & j & j & j \\
\end{array}$$

Now, all we have to do is find $g(x)$. First, I'll simplify the function.

$$g(x) = -3(x-3) + 1$$

$$g(x) = -3x + 9 + 1$$

$$g(x) = -3x + 10$$

I don't need to go any further because then I'm dealing with fractions which may over-complicate things. Let's see how it goes. 

Obviously, it'll be easiest to start with $x = 0$

$$g(0) = -3(0) + 10$$

$$g(0) = 10$$

We have at least one value of $g(x)$ we can populate:

$$\begin{array}{c|c|c|c}
    x    & -1 & 0 & 1 \\ \hline
    g(x) & j & 10 & j \\
\end{array}$$

And, $g(-1)$:

$$g(-1) = -3(-1) + 10$$

$$g(-1) = 3 + 10$$

$$g(-1) = 13$$

And, $g(1)$:

$$g(1) = -3(1) + 10$$

$$g(1) = -3 + 10$$

$$g(1) = 7$$

$$\begin{array}{c|c|c|c}
    x    & -1 & 0 & 1 \\ \hline
    g(x) & 13 & 10 & 7 \\
\end{array}$$

We'll do the same with the equation:

$$h(t) = -(t + 2) -2$$

First, I'll simplify:

$$h(t) = -t - 2 - 2$$

$$h(t) = -t -4$$

I'll start using the t values of -1, 0 and 1:

$$\begin{array}{c|c|c|c}
    t    & i & i & i \\ \hline
    h(t) & j & j & j \\
\end{array}$$

$t = -1$:

$$h(-1) = -(-1) - 4$$

$$h(-1) = -3$$

$t = 0$:

$$h(0) = -(0) - 4$$

$$h(0) = -4$$

$t = 1$:

$$h(1) = -(1) - 4$$

$$h(1) = -5$$

And our final results:
$$\begin{array}{c|c|c|c}
    x    & -1 & 0 & 1 \\ \hline
    g(x) & -3 & -4 & -5 \\
\end{array}$$

## Functions and Equations

###For a given input value $x$, the function $g$ outputs a value of $y$ to satisfy the following equation:

$$-4x - 6 = -5y + 2$$

Write a formula for $g(x)$ in terms of $x$.

####Solution

We're being asked to find the formula from the above equation that will take $g(x)$ and produce that equation. In that $x$ is the parameter we are taking as input, we need to find, or isolate, $y$.

$$-4x - 6 = -5y + 2$$

$$5y = 4x + 8$$

$$y = \frac{4x}{5} + {8}{5}$$

And our formula is:

$$g(x) = \frac{4}{5}x + \frac{8}{5}$$

###For a given input value $q$, the function $g$ outputs a value $r$ to satisfy the following equation:

$$-7q + 12r = 3q - 4r$$

Write a formula for $g(q)$ in terms of $q$.

####Solution

So, here we're taking input of $q$ to find the value of $r$. Therefore, we need to isolate $r$.

$$-7q + 12r = 3q - 4r$$

$$12r + 4r = 3q + 7q$$

$$16r = 10q$$

$$r = \frac{10}{16}q$$

$$r = \frac{5}{8}q$$

So our final function is:

$$g(q) = \frac{5}{8}q$$
