---
# General format taken from <https://dev.to/yelluw/how-to-do-technical-blogging>

# Title
#   - Short title, start with verb, provide brief explanation what this is about.
#   - Examples:
#     - Writing a $technology $thing
#     - Developing with $technology
#     - Testing $techonology with $thing
title: SQL Template

# Summary
#   - 2-3 sentences. 
#   - Summarize what post is about.
#   - Should be closely related to title
#   - Should be smooth transition from title
summary: > 
  Use this template for SQL technical posts. 

# Description is not used at this time but may be implemented later.
description: NULL

# Abstract is not used at this time but may be implemented later.
abstract: NULL

author: Tim Trice
date: '2017-11-04'

# Slug
#   - Use title without conjunctions and determiners.
#   - All lower-case.
slug: slug

# Categories
#   - General classification of post.
#     - SQL
#     - R
#     - Python
categories:
  - SQL

# Tags
#   - Calls specific to category such as functions, or classes.
#     - In SQL, "SELECT", "ORDER BY", "PARTITION", etc.
#     - In R, "dplyr", "rattle", "vector", etc."
tags:
  - SELECT

output:
  blogdown::html_page:
    toc: true
    toc_depth: 6
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo        =    TRUE, 
                      error       =    TRUE, 
                      max.print   =    NA, 
                      connection  =    "con")
```

```{r libraries, include = FALSE}
library(RMariaDB)                                #todo [Use correct db library]
```

```{r db-con-params, include = FALSE}
host     <-   Sys.getenv("mariadb.host")         #todo [Confirm host]
user     <-   Sys.getenv("mariadb.user")         #todo [Confirm user]
password <-   Sys.getenv("mariadb.password")     #todo [Confirm password]
driver   <-   MariaDB()                          #todo [Confirm driver]
```

```{r db-con, include = FALSE}
con <- dbConnect(driver, 
                 host = host,
                 user = user, 
                 password = password)
```

### Requirements

  * What are the requirements for reproducing the contents of this post? 

  * Include versions of all called packages, as well.

  * Pretend you found this article via Google and need to know if it pertains 
    to your issue.

Example: 

Python 3.6.1

Text Editor (Vim was used for the examples)

Operating System (Linux, MacOS, Windows) with version information (Ubuntu 16, MacOS 10.12, Windows 10 Professional)

- OR - 

```{sql db-version, tab.cap = "Server Info"}
SELECT @@version_comment AS `Version`, 
  @@version_compile_machine AS `Version Compile Machine`, 
  @@innodb_version AS `InnoDB Version`, 
  @@version_compile_os AS `Version Compile OS`;
```

### Introduction

  * What is it you will cover in your post? 

  * Do not be technical in this piece

Python has become the defacto glue language in software development. In it's newest version (3.6.1), the language has evolved into a powerful tool that is easy to pick up. Python is included as a default in most Linux distros, and in MacOS. Such widespread adoption means you will probably not need to install anything in order to run Python on your machine. Follow along as I explain how to use Python's F-Strings in a Linux environment.

### Multiple step explanation

  * Break everything down to the smallest step possible. 
  
  * Stay specific to the step
  
  * Provide visual clues where at all possible.
  
  * Readers will look for the deepest scope first, then work backwards.

### Examples

  * Provide examples that directly correlate to the explanations above. 
  
  * Range examples from very simple to rather complex. 

### Tips #todo [Remove this section - this is for the author]

  * When in doubt, explain (but carefully and organized). Use visual cues when possible. 
  
  * Keep it simple. 
  
  * Look at [Stripe](https://stripe.com/docs/connect/accounts) documentation for examples.
  
  * Run code within the article. 

### Summary

  * Add links to sources.
  
  * Thank those who helped contribute.
  
  * Close with direct contact info.

```{r close-db, include = FALSE}
dbDisconnect(con)
```
