---
wp_title: Geometry
wp_draft: TRUE
wp_categories: Geometry
wp_mt_keywords: geometry
---

```{r, echo = FALSE}
library(ggplot2)
```

## Geometry

### Area and Perimeter

$$A = \left(1/2\right)bh$$ 

```{r, fig.cap = "Sample"}
df <- data.frame(x = c(1, 5, 5), 
                 y = c(2, 2, 5))

ggplot(data = df, aes(x = x, y = y)) + 
    geom_polygon(fill = "#0000FF", alpha = 0.1) + 
    geom_text(label = "h", x = 4.8, y = 3.5) + 
    geom_text(label = "b", x = 3, y = 2.1)
```

Above, we have a pretty simple right triangle. To find $A$ of the triangle we first need $h$ which is the height of the triangle or $5-2$. Then the base $b$ which is the width or $5-1$. 

$$A = \left(1/2\right)\left(4\right)\left(3\right)$$

$$A = \left(1/2\right)\left(12\right)$$

$$A = 6$$

A slight more complex example would be an angled right triangle such as below:

```{r}
df <- data.frame(x = c(3, 7, 1), 
                 y = c(1, 2, 5))

ggplot(data = df, aes(x = x, y = y)) + 
    geom_polygon(fill = "#0000FF", alpha = 0.1) + 
    geom_text(label = "h", x = 2.2, y = 3) + 
    geom_text(label = "b", x = 4.6, y = 1.6)
```

Because of the angle we can't just find $h$ by substracting 1 from 5. We can, however, build right triangles around the original triangle, do some additional tricks and get the area of the original triangle.

```{r}

df <- data.frame(x = c(3, 7, 1, 1, 3, 1, 1, 7, 7, 3, 7, 7), 
                 y = c(1, 2, 5, 1, 1, 5, 5, 5, 2, 1, 1, 2), 
                 z = c(rep("A", 3), rep("B", 3), rep("C", 3), rep("D", 3)))
ggplot(data = df, aes(x = x, y = y, fill = z, colour = z, group = z)) + 
    geom_polygon(aes(alpha = 0.1))
```

So, now our original triangle, $\triangle A$ is surrounded by new triangles $\triangle B$, $\triangle C$, and $\triangle D$. From the broader standpoint, we've created a rectangle which I will name $\Box E$.

First, I'll calculate $A$ of $\Box E$ which is six units wide and 4 units high.

$$A of \Box E = wl$$

$$A of \Box E = \left(6\right)\left(4\right)$$

$$A of \Box E = 24$$

Next, we can calculate the area of each $\triangle B$, $\triangle C$ and $\triangle D$:

$$A of \triangle B = \left(1/2\right)\left(2\right)\left(4\right)$$

$$A of \triangle B = 4$$

$$A of \triangle C = \left(1/2\right)\left(6\right)\left(3\right)$$

$$A of \triangle C = 9$$

$$A of \triangle D = \left(1/2\right)\left(4\right)\left(1\right)$$

$$A of \triangle D = 2$$

To find the area of $\triangle A$ we subtract the sum of the areas of $\triangle B$, $\triangle C$ and $\triangle D$ from $\Box E$:

$$A of \triangle A = \Box E - \triangle B - \triangle C - \triangle D$$

$$A of \triangle A = 24 - 4 - 9 - 2$$

$$A of \triangle A = 9$$


